A Formal Verification Framework for Smart Contracts Based on Symbolic Execution

### 2024-01-17 papers/static-checker/ecoop2024/ecoop2024.tex
kolibri code and picture
code updated
cosmetics

\subsection{Simple contracts}
\label{sec:simple-contracts}
Listing~\ref{lst:smart-contract-property-specification} shows the
syntax of our property specification language for a simple contract. Here, \KMCONTRACT,
\KSPEC, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are
keywords. The \KNAME\ is replaced by the contract name. In the input
and output fields, the input and output terms are specified. The input and output terms represent Michelson terms, which can be either concrete values or symbolic values, including variables. The terms are specified using the syntax defined in Fig. \ref{fig:syntax-terms-predicates}. A term \textbf{t} comprises constants of the sort \text{nat} (natural number), \text{int} (integer), \text{string} (string), and \text{b} (bytes) along with their respective types. As a syntactic sugar, the type can be omitted except for the type \text{nat} to distinguish it from integer numbers. It also encompasses all blockchain constants \textbf{c}, such as Amount and Sender. When a variable appears for the first time, it must be accompanied by its corresponding type declaration. Subsequent mentions can omit the type declaration. The language supports essential Michelson types, as shown in Fig. \ref{fig:types}, where \textbf{cpt} denotes comparable types. The types include basic data types such as integer, string, list, set, pair, option and map. Additionally, there are some typical types related to the blockchain, such as \text{mutez} and \text{contract T}, representing types for tokens and smart contracts on the Tezos blockchain, respectively.

Since Michelson operates on a stack, we assume that the stack has only one
element, as specified in the input term. Similarly, the output
stack is assumed to be a stack with only one element, which is
specified in the output field. The syntax allows us to specify both
incomplete and complete smart contracts\footnote{The tool has been adapted to accommodate incomplete contract codes that take input stacks and produce output stacks with more than one element. This modification is feasible from a technical standpoint, as the current tool can effectively handle such cases.}. For a complete smart contract, the output should be a pair consisting of an operation list and the storage's state after the completion of the code.  For verification purposes we elide the operation list and represent the output solely by the term of the value in the storage after the execution concludes. In some cases, the output term can be omitted for a complete contract, given that the input term must be a pair
of parameters and storage. Hence the storage term is part of the
input term. In such cases, the prefix \lstinline/post_/ is used
before the variables from the input term to indicate the corresponding
storage value after the execution.

Formally, the pre- and postcondition instances are formulas in first-order logic that can be handled by an SMT solver (in our case, Z3). The specification language is constituted of predicates.  The syntax predicates are defined in Figure~\ref{fig:syntax-terms-predicates}. Formulas that are used in assertions or added to solvers are terms of sort Boolean. Otherwise, terms of
Boolean and non-Boolean sort may be mixed in any combination where
sorts match up. Several functions are defined, such as the function \text{get\_map (e1, e2)} for accessing an element in a map. This function returns \text{None} if there is no key \text{e1} in the map \text{e1}, or the corresponding value of the key otherwise. Another function, \text{get\_contract (e1, e2)}, casts the address \text{e2} to a real contract on the Tezos blockchain. It returns \text{None} if there is no existing contract on the Tezos block chain with an entrypoint of type \text{e1} associated with the address \text{e2}, or it returns the contract otherwise.

Some other functions are defined as predicates, such as \text{transfer\_token} (\text{e1}, \text{e2}, \text{e3}). This implies that there is an operation in the operation list, after executing the smart contract code, which indicates the transfer of \text{e2} amount of tokens to the address \text{e3} along with the data \text{e1}\footnote{While some additional functions and both universal and existential quantifiers are implemented, they pose no technical problems and are similar to existing supported functions. However, the conversion to Z3 formulas in the Z3 OCaml library takes some time.}.

@@

Let us consider a fragment code of the contract \texttt{Sub}  in Listing
\ref{lst:sub-contract-specification} as an example of a property
specification.  Here, the program takes as input a stack with only one
element, which is a pair of two integers, symbolically specified in
the \lstinline|input| field. The program first duplicates the pair. It then unpairs the input pair to obtain two elements as two integers \lstinline/x/ and \lstinline/y/.
Next, the program performs a comparison between
these two integers. If \lstinline/x/ is greater than or equal to
\lstinline/y/, the program constructs a new stack with a solitary
element at the top representing the subtract of  \lstinline/x/ and
\lstinline/y/, designated as \lstinline/z/ by the \lstinline|output| directive. Otherwise, the execution
terminates with an error message containing the content
\lstinline|'Unexpected pair'|.

@@

To illustrate entrypoint relations, let us consider a smart contract
that models an online auction. The contract has two entrypoints,
\lstinline|%bid| and \lstinline|%close|, which serve as functions for
bidding and closing the contract, respectively.  The smart contract
storage is a pair containing as its first element a boolean value
\lstinline|auction_open| that indicates whether the contract is still
open for bidding or already closed. The second element is another
nested pair, where the first element stores the highest bidder's
address, and the second element is the contract owner's address. The
current highest bid is reflected in the balance of the contract.

@@

There are several properties that we want to guarantee for this
contract. Let us have a look at the specification for the auction
contract in Listing \ref{lst:auction-contract-specification}. We omit
the Michelson code of the smart contract due to its length.  For the
\lstinline/%bid/ entrypoint, the input stack has only one element,
which is a pair consisting of the parameter and the current
storage. The parameter is \lstinline/Unit/ of type \lstinline/unit/.
The current storage is represented as a nested pair with the
structure described by the \lstinline|storage| phrase. 
The output term should maintain the required term of the storage,
but the values may be changed and these changes are recorded in
different symbolic variables. In this specification, we add the prefix
\lstinline/post_/ to the variable names to refer to the storage
returned by the program. The output term could be safely removed in
this example because the tool automatically generates an output
term corresponding the one specified here.

@@

Let us return to the auction contract. It is important to guarantee
that contract deployment sets the  \lstinline/auction_open/ flag to
\TRUE, indicating that the contract is open for bidding, i.e., users
should be able to bid on the contract or the contract owner should be
able to close it (lines 28-35). Furthermore, the bid call can happen
successfully after another if the new bid is higher than the previous
one (line 36). The  \lstinline/%close/ entrypoint can be called
anytime if the caller is the contract owner (line 34). However, it
should not be possible to call the  \lstinline/%bid/ entrypoint after
the auction is closed (line 42), and, of course, it cannot be closed
again if it is already closed (line 43). This sequence  illustrates
the life cycle of the auction contract, which is depicted in Figure
\ref{fig:auction-life-cycle}.

>>>>

\subsection{Simple contracts}
\label{sec:simple-contracts}
Listing~\ref{lst:smart-contract-property-specification} shows the
syntax of our property specification language for a simple contract. Here, \KMCONTRACT,
\KSPEC, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are
keywords. The \KNAME\ is replaced by the contract name. The \KINPUT{}
and \KOUTPUT{} fields specify the input and output terms, which are
patterns on Michelson values that correspond to the top value of the
initial and final stack. A pattern can be a concrete value or it may
contain symbolic variables. Fig.~\ref{fig:syntax-terms-predicates}
contains the grammar of terms: A term \textbf{t} can consist of
constants of the sort \text{nat} (natural number), \text{int}
(integer), \text{string} (string), and \text{b} (bytes) along with
their respective types. The type annotation may be omitted except for
the type \text{nat} to distinguish it from integer numbers.  It also
encompasses all blockchain constants \textbf{c}, such as Amount and
Sender. When a variable appears for the first time, it must be
accompanied by its corresponding type declaration. Subsequent mentions
can omit the type declaration. The language supports essential
Michelson types, as shown in Fig.~\ref{fig:types}, where \textbf{cpt}
denotes comparable types. Types include basic data types such as
integer, string, list, set, pair, option, and map. Additionally, there
are some typical types related to the blockchain, such as \text{mutez}
and \text{contract T}, representing types for tokens and smart
contracts with a parameter of type $T$ on the Tezos blockchain.

Since Michelson operates on a stack, we assume that the stack has only one
element, as specified in the input term. Similarly, the output
stack is assumed to be a stack with only one element, which is
specified in the output field. The syntax allows us to specify both
incomplete and complete smart contracts\footnote{The tool can
  accommodate incomplete contract codes that take input stacks and
  produce output stacks with more than one element, but this is
  currently not reflected in the syntax.}. For a complete smart
contract, the output should be a pair consisting of an operation list
and the storage's state after the completion of the code.  For
verification purposes we elide the operation list and represent the
output solely by the term of the value in the storage after the
execution concludes. In some cases, the output term can be omitted for
a complete contract, given that the input term must be a pair 
of parameters and storage. Hence the storage term is part of the
input term. In such cases, the prefix \lstinline/post_/ is used
before the variables from the input term to indicate the corresponding
storage value after the execution.

Formally, the pre- and postcondition instances are formulas in
first-order logic that can be handled by an SMT solver (in our case,
Z3). The specification language is built from
predicates. Figure~\ref{fig:syntax-terms-predicates} defines the
syntax of predicates. Formulas that are used in assertions or added to solvers are terms of sort Boolean. Otherwise, terms of
Boolean and non-Boolean sort may be mixed in any combination where
sorts match up. Several functions are predefined, such as the function
\text{get\_map (e1, e2)} for accessing an element in a map. This
function returns \text{None} if there is no key \text{e1} in the map
\text{e1}, or the corresponding value of the key otherwise. Another
function, \text{get\_contract (e1, e2)}, casts the address \text{e2}
to a real contract on the Tezos blockchain. It returns \text{None} if
there is no existing contract on the Tezos block chain with an
entrypoint of type \text{e1} associated with the address \text{e2}, or
it returns the contract otherwise.

Some other functions are defined as predicates, such as
\text{transfer\_token} (\text{e1}, \text{e2}, \text{e3}). This
predicate holds if there is an operation in the operation list, after
executing the smart contract code, which indicates the transfer of
\text{e2} amount of tokens to the address \text{e3} along with the
parameter \text{e1}\footnote{It is straightforward to add further
  predicates if a mapping to Z3 formulas can be defined.}.

@@

Let us consider a code fragment of the contract \texttt{Sub}  in Listing
\ref{lst:sub-contract-specification} as an example of a property
specification.  Here, the program takes as input a stack with only one
element, which is a pair of two integers, symbolically specified in
the \lstinline|input| field. The program first duplicates the pair. It
then unpairs the input pair to obtain two elements as two integers
\lstinline/x/ and \lstinline/y/. 
Next, the program performs a comparison between
these two integers. If \lstinline/x/ is greater than or equal to
\lstinline/y/, the program constructs a new stack with a solitary
element at the top representing the subtract of  \lstinline/x/ and
\lstinline/y/, designated as \lstinline/z/ by the \lstinline|output| directive. Otherwise, the execution
terminates with an error message containing the content
\lstinline|'Unexpected pair'|.

@@

To illustrate entrypoint relations, let us consider a smart contract
that models an online auction. The contract has two entrypoints,
\lstinline|%bid| and \lstinline|%close|, which serve as functions for
bidding and closing the contract, respectively.  The smart contract
storage is a pair containing as its first element a boolean value
\lstinline|auction_open| that indicates whether the contract is still
open for bidding or already closed. The second element is another
nested pair, where the first element stores the highest bidder's
address, and the second element is the contract owner's address. The
current highest bid is reflected in the balance of the contract before
the next bid is invoked.

@@

There are several properties that we want to guarantee for this
contract. Let us have a look at the specification for the auction
contract in Listing \ref{lst:auction-contract-specification}. We omit
the Michelson code of the smart contract due to its length.  For the
\lstinline/%bid/ entrypoint, the input stack has only one element,
which is a pair consisting of the parameter and the current
storage. The parameter is \lstinline/Unit/ of type \lstinline/unit/.
The current storage is represented as a nested pair with the
structure described by the \lstinline|storage| phrase. 
The output term should maintain the required term of the storage,
but the values may be changed and these changes are recorded in
different symbolic variables. In this specification, we add the prefix
\lstinline/post_/ to the variable names to refer to the storage
returned by the program.
% The output term could be safely removed in
% this example because the tool automatically generates an output
% term corresponding the one specified here.

@@

Let us return to the auction contract. It is important to guarantee
that contract deployment sets the  \lstinline/auction_open/ flag to
\TRUE, indicating that the contract is open for bidding, i.e., users
should be able to bid on the contract or the contract owner should be
able to close it (lines 28-35). Otherwise, the owner's original
deposit is lost. Furthermore, the bid call can happen
successfully after another if the new bid is higher than the previous
one (line 36). The  \lstinline/%close/ entrypoint can be called
anytime if the caller is the contract owner (line 34). However, it
should not be possible to call the  \lstinline/%bid/ entrypoint after
the auction is closed (line 42), and, of course, it cannot be closed
again if it is already closed (line 43). This sequence  illustrates
the life cycle of the auction contract, which is depicted in Figure
\ref{fig:auction-life-cycle}.
